Novel improvements in the Material Point Method to face dynamic problems: Local-maximum entropy approximation and an explicit predictor-corrector scheme. \thanks{Funding: This study was funded by Agustín de Betancourt Foundation (grant number 262390106114).}

### 2020-03-06 main.tex
Cambios 6-3-2020

Finally, to solve the equation \eqref{eq:particle_balance_forces3}, a second order
temporal integration scheme is required. Therefore, time is
discretized in to a finite set of time steps $k = 1\ldots ,Nt$, where $k$ is the current time step and $N_t$
is the total number of time steps. Once the nodal equilibrium equation it is solved, the values in the nodes are
interpolated back in to the particles and each particle it is advected
to the new position,
\begin{equation}
  \label{eq:Updated_Lagrangian}
  \dot{\vec{v}}_p = \sum^{N_n}_{I}N_{Ip}\ \vec{a}_{I},\quad and\quad
  \dot{\vec{x}}_{p} = \sum^{N_n}_{I}N_{Ip}\ \vec{v}_{I}  
\end{equation}
In the MPM literature, this equations
(\ref{eq:particle_balance_forces3}) and (\ref{eq:Updated_Lagrangian}),
are solved with an explicit forward Euler algorithm. In the next
section, an alternative time integration scheme based on the Newmark a-form is presented.

@@

The time integration scheme of the MPM has been described in detail by many researchers
\cite{Sulsky1994}, \cite{Bardenhagen2002}, \cite{thesis_Andersen_2009} and summarized
in Figure \ref{fig:MPM_algorithm}. Other authors have proposed many
others time integration alternatives like
\cite{Guilkey2003}, \cite{Tran2019e}, \cite{Charlton2017}. In the first publication on
the MPM \cite{Sulsky1994}, the nodal acceleration was employed to
update the particles as
\begin{equation}
  \label{eq:Sulsky-1994-UL-v}
  \vect{v}_p^{k+1} = \vect{v}_p^{k} + \Delta t\
  \sum^{N_n}_{I}N_{Ip}^{k}\ \vec{a}_{I}^{k}
\end{equation}
\begin{equation}
  \label{eq:Sulsky-1994-UL-x}
  \vect{x}_p^{k+1} = \vect{x}_p^{k} + \Delta t\ \sum^{N_n}_{I}N_{Ip}^{k}\ \vec{v}_{I}^{k}.
\end{equation}
However, as Andersen (2009)\cite{thesis_Andersen_2009} point out, this algorithm has been shown to be numerically unstable due to that
$\vect{f}_I^{int,k}$ can be finite for an infinitesimal nodal mass
$\tens{m}$. This can lead to numerical issues when nodal acceleration
is obtained for evaluating \eqref{eq:Sulsky-1994-UL-x},\eqref{eq:Sulsky-1994-UL-v}. Hence, a
corrected version of this algorithm in shown in Zhang {\it et al.}
(2016)\cite{Zhang_book_2016}
\begin{equation}
  \label{eq:Zhang-2016-UL-x}
  \vect{x}_p^{k+1} = \vect{x}_p^{k} + \Delta t\ \sum^{N_n}_{I} \frac{N_{Ip}^{k}\ \vec{p}_{I}^{k}}{\tens{m}_I}.  
\end{equation}
\begin{equation}
  \label{eq:Zhang-2016-UL-v}
  \vect{v}_p^{k+1} = \vect{v}_p^{k} + \Delta t\
  \sum^{N_n}_{I}\frac{N_{Ip}^{k}\ \vec{f}_{I}^{k}}{\tens{m}_I},
\end{equation}
Later Tran \& Solowski (2019)\cite{Tran2019e} presented a
generalized-$\alpha$ scheme for the MPM inspired in the explicit time
integration algorithm proposed by Chung \& Hulbert
(1993)\cite{Geranlized_alpha_1993}, but with the particularity that
the acceleration is evaluated both in the beginning and the end of the
time step.
\begin{equation}
  \label{eq:Tran-2019-GA-v}
  \vect{v}_p^{k+1} = \vect{v}_p^{k} + \Delta t\
  \sum^{N_n}_{I} N_{Ip}^{k}\ \left[(1 - \gamma)\ \vect{a}_I^{k} +
    \gamma\ \vect{a}_I^{k+1} \right],\\
\end{equation}
\begin{equation}
\label{eq:Tran-2019-GA-x}
  \vect{x}_p^{k+1} = \vect{x}_p^{k} + 
  \sum^{N_n}_{I}N_{Ip}^{k} \left[ \Delta t\ \vec{v}_{I}^{k}+ \Delta t^2\left( (\frac{1}{2} - \beta)\
    \vec{a}_{I}^{k} + \beta\ \vec{a}_{I}^{k+1} \right) \right]
\end{equation}
\begin{equation}
  \label{eq:Tran-2019-GA-a}
  \vect{a}_p^{k+1} = \sum^{N_n}_{I}N_{Ip}^{k}\ \vec{a}_{I}^{k+1}.
\end{equation}

This scheme has prof to damps out the higher frequency noises
\cite{Tran2019e}. But it can present the same numerical instabilities
as in \eqref{eq:Sulsky-1994-UL-x},\eqref{eq:Sulsky-1994-UL-v} when nodal masses become infinitesimal.
Here we propose a explicit predictor-corrector scheme for MPM based on the Newmark a-form with
$\gamma = 0.5$ and $\beta = 0$ which is the central difference
explicit. This method is devoted to solve a system of equations of type
\begin{equation*}
  \Matrix{M}_{IJ}\ddot{\Vector{d}}_{J} + \Matrix{C}_{IJ}\dot{\Vector{d}}_{J} +
  \Matrix{K}_{IJ}\Vector{d}_{J} = \Vector{F}_{I}
\end{equation*}
Nevertheless it can be applied for an isolated mass, therefore it is
possible to apply this methods successfully in the MPM framework as
was proved by \cite{Tran2019e}. Taking the predictor
definition from the classic literature \cite{Hughes2000} and
calculating particles velocity and acceleration as described in 
\cite{Zhang_book_2016}, we get the following expression for the
\textbf{predictor} stage for both velocity and displacement.
\begin{equation}
  \label{eq:5}
  \vec{x}_p^{k+1} = \vec{x}_p^n + \Delta t\
  \underbrace{\sum^{N_n}_{I}\frac{N_{Ip}^k\ \vec{p}_{I}^{k}}{\tens{m}_I^k}}_{\vec{v}_p^{k}} + \frac{1}{2}\Delta
  t^2\ \underbrace{\sum^{N_n}_{I}\frac{N_{Ip}^k\ \vec{f}_{I}^{k}}{\tens{m}_I^k}}_{\vec{a}_p^{k}}
\end{equation}
\begin{equation}
  \label{eq:Predictor-velocity}
  \vec{v}_p^{k+1} = \vec{v}_p^n + (1 - \gamma)\ \Delta t\
  \underbrace{\sum^{N_n}_{I}\frac{N_{Ip}^k\ \vec{f}_{I}^{k}}{\tens{m}_I^k}}_{\vec{a}_p^{k}}
\end{equation}

Consequently, the \textbf{corrector} stage for the particles velocity is in the
following way
\begin{equation}
  \vec{v}_{p}^{k+1} = \vec{v}_{p}^{k+1} +
  \gamma\ \Delta t\ \underbrace{\sum^{N_n}_{I}\frac{N_{Ip}^{k+1}\ \vec{f}_{I}^{k+1}}{\tens{m}_I^{k+1}}}_{\vec{a}_p^{k+1}}  
\end{equation}

Notice that the predictor of the particle displacement is
computed using information only from the predictor of the
previous step. However, velocities are computed using half information
of the past corrector and half of the past predictor. Therefore here we share
similarities with the \textbf{leapfrog integration} which updates the
position at full time step, but updates the velocity at half time
steps. Due to its simplicity allows be implemented with minor modifications
over a standard forward Euler. It is summarized in shape of
pseudo-algorithm \ref{alg:PCE-algorithm}.

\begin{algorithm}
  \label{alg:PCE-algorithm}
  \floatname{algorithm}{Algorithm}
  \renewcommand{\thealgorithm}{}
  \caption{Explicit scheme : Newmark central difference}
  \begin{algorithmic}[1]
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% º
    \STATE \textbf{Explicit Newmark Predictor} ($\gamma = 0.5$) :
    \begin{align*}
      &\vec{x}_p^{k+1} = \vec{x}_p^n + \Delta t\
        \sum^{N_n}_{I}\frac{N_{Ip}^k\ \vec{p}_{I}^{k}}{\tens{M}_I^k} +
        \frac{1}{2}\Delta t^2\ \sum^{N_n}_{I}\frac{N_{Ip}^k\ \vec{f}_{I}^{k}}{\tens{M}_I^k}\\
      &\vec{v}_p^{k+1} = \vec{v}_p^n + (1 - \gamma)\
        \sum^{N_n}_{I}\frac{N_{Ip}^k\ \vec{f}_{I}^{k}}{\tens{M}_I^k}\Delta t        
    \end{align*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Discard the previous nodal values}.
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Transference of particles kinetics information to the
      mesh}:\\
    Calculate the lumped mass matrix $\tens{M}_{I}$,
    and the nodal momentum $\vec{p}_{I}$ with the predicted values.
    \begin{equation*}
      \tens{M}_{I} = \sum_{p}^{N_p}N_{Ip}^{k+1}\ m_p,
    \end{equation*}
    \begin{equation*}
      \vec{p}_{I}^{k+1} = \sum_{p}^{N_p} N_{Ip}^{k+1}\ m_p\ \vec{v}_{p}^{k+1},
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Impose essential boundary conditions over momentum nodal
      values}:\\
    At the fixed boundary, set $\vec{p}_{I}^{k+1} = 0$. Later, get the
    nodal values of the velocity field $\vec{v}_{I}^{k+1} =
    \frac{\vec{p}_{I}^{k+1}}{\tens{M}_{I}}$
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Deformation tensor increment calculation}.
    \begin{align*}
      &\dot{\tens{\varepsilon}_{p}}^{k+1} = \frac{1}{2}\ \sum_{I}^{N_n}
        \Grad{N_{Ip}^{k+1}}\vec{v}_{I}^{k+1} +
        (\Grad{N_{Ip}^{k+1}}\vec{v}_{I}^{k+1})^T \\
      &\Delta \tens{\varepsilon}_{p}^{k+1} = \Delta t\ \dot{\tens{\varepsilon}_{p}}^{k+1}
    \end{align*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Update the density field}:\\
    \begin{equation*}
      \rho_p^{k+1} = \frac{\rho_p^k}{1 + \mathit{tra}\left[\Delta\tens{\varepsilon}_{p}^{k+1}\right]}.
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Balance of forces calculation}:\\
    Calculate the total grid nodal force $\vec{f}_{I}^{k+1} =
    \vec{f}_{I}^{int,k+1} + \vec{f}_{I}^{ext,k+1}$ evaluating
    \eqref{eq:nodal_internal_forces} and \eqref{eq:nodal_external_forces} in the time step $k+1$. In the
    grid node $I$ is fixed in one of the spatial dimensions, set it to
    zero to make the grid nodal acceleration zero in that direction.\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Integrate the grid nodal momentum equation :}\\
    Calculate the nodal momentum $\vec{p}_{I}$ in $k+1$. 
    \begin{equation*}
      \vec{p}_{I}^{k+1} = \vec{p}_{I}^{k} + \vec{f}_{I}^{k+1}\Delta t
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Explicit Newmark Corrector}:
    \begin{equation*}
      \vec{v}_{p}^{k+1} = \vec{v}_{p}^{k+1} +
      \gamma \sum^{N_n}_{I}\frac{N_{Ip}^{k+1}\ \vec{f}_{I}^{k+1}}{\tens{m}_I^{k+1}}\Delta t  
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
  \end{algorithmic}
\end{algorithm}

\begin{figure}\sidecaption
  \centering
  \resizebox{0.6\hsize}{!}{
    \includegraphics[width=\textwidth]{./Figures/MPM_scheme}
  }
  \caption{MPM standard algorithm.}
  \label{fig:MPM_algorithm}
\end{figure}

@@

\section{Numerical results : 1D Elastic bar}
\label{sec:dynamic-benchmark}

In this section, the benchmark proposed by Dyka \& Ingel
(1995)\cite{Dyka1995} is considered to illustrate the
capacity of MPM \cite{Sulsky1994} and OTM
models to avoid tensile instabilities. Here we employ our in-house
codes for both MPM and OTM simulations.

In the one-dimensional bar sketched in the figure \ref{fig:Dyka_Bar},
the left end of the bar is fixed and the right and an initial velocity
$v_0 = 5\ m/s$ is given to the last quarter of it in the x positive
direction. The length of it is 0.1333 meters with an unit section. The
elastic parameters consider for this test are:
\begin{itemize} 
\item  Density : $7833\ kg/m3$
\item  Poisson ratio : $0$
\item  Elastic modulus : $200 \cdot 10^9\ Pa$
\end{itemize}

The total simulation time is 0.0002 seconds, a direct consequence of
that is we let the elastic wave generated travel thorough the bar
(from A to C and back to A) at least four times. For the spatial
discretization we will set five different values of the it : 0.1,
0.05, 0.01, 0.005 and 0.001 millimeters. The initial mesh will be
composed by a single particle in the center of the element. Regarding
this, here we give one consideration to properly modellize this
benchmark under the MPM framework. Notice that the free border of the
bar has a horizontal displacement of 0.033 millimeters, therefore a
computational domain with an extra gap of 0.033 millimeters is required in
order to accommodate the unconstrained displacement of the particles in the
left border of the bar. Here below the time step is controlled by the courant number, in
this case, we adopt a middle value of it (CFL=0.5) as a compromise
between stability and numerical accuracy.

The analytical solution in terms of velocity
\ref{fig:vel_analytics_dyka}, and stress
\ref{fig:stress_analytics_dyka} can be found in the appendix
\ref{app:analytical_sol} given by the Method of characteristics. For
the convergence analysis, the root-mean-square (RMS) error in the
displacement is computed. RMS error is defined as follows

@@

>>>>

Finally, to solve the equation \eqref{eq:particle_balance_forces3}, a second order
temporal integration scheme is required. Therefore, time is
discretized in to a finite set of time steps $k = 1\ldots ,Nt$, where $k$ is the current time step and $N_t$
is the total number of time steps. Once the nodal equilibrium equation it is solved, the values in the nodes are
interpolated back in to the particles and each particle it is advected
to the new position,
\begin{equation}
  \label{eq:Updated_Lagrangian}
  \dot{\vec{v}}_p = \sum^{N_n}_{I}N_{Ip}\ \vec{a}_{I},\quad and\quad
  \dot{\vec{x}}_{p} = \sum^{N_n}_{I}N_{Ip}\ \vec{v}_{I}  
\end{equation}
In the MPM literature, this equations
(\ref{eq:particle_balance_forces3}) and (\ref{eq:Updated_Lagrangian}),
are solved with an explicit forward Euler algorithm. The time integration scheme of the MPM has been described in detail by many researchers
\cite{Sulsky1994}, \cite{Bardenhagen2002}, \cite{thesis_Andersen_2009} and summarized
in Figure \ref{fig:MPM_algorithm}.

\begin{figure}\sidecaption
  \centering
  \resizebox{0.55\hsize}{!}{
    \includegraphics[width=\textwidth]{./Figures/MPM_scheme}
  }
  \caption{MPM standard algorithm.}
  \label{fig:MPM_algorithm}
\end{figure}

Other authors have proposed many others time integration alternatives
like \cite{Guilkey2003}, \cite{Tran2019e}, \cite{Charlton2017}. In the
first publication on the MPM \cite{Sulsky1994}, the nodal acceleration
was employed to update the particles as
\begin{equation}
  \label{eq:Sulsky-1994-UL-v}
  \vect{v}_p^{k+1} = \vect{v}_p^{k} + \Delta t\
  \sum^{N_n}_{I}N_{Ip}^{k}\ \vec{a}_{I}^{k}
\end{equation}
\begin{equation}
  \label{eq:Sulsky-1994-UL-x}
  \vect{x}_p^{k+1} = \vect{x}_p^{k} + \Delta t\ \sum^{N_n}_{I}N_{Ip}^{k}\ \vec{v}_{I}^{k}.
\end{equation}
However, as Andersen (2009)\cite{thesis_Andersen_2009} point out, this algorithm has been shown to be numerically unstable due to that
$\vect{f}_I^{int,k}$ can be finite for an infinitesimal nodal mass
$\tens{m}$. This can lead to numerical issues when nodal acceleration
is obtained for evaluating \eqref{eq:Sulsky-1994-UL-x},\eqref{eq:Sulsky-1994-UL-v}. Hence, a
corrected version of this algorithm in shown in Zhang {\it et al.}
(2016)\cite{Zhang_book_2016}
\begin{equation}
  \label{eq:Zhang-2016-UL-x}
  \vect{x}_p^{k+1} = \vect{x}_p^{k} + \Delta t\ \sum^{N_n}_{I} \frac{N_{Ip}^{k}\ \vec{p}_{I}^{k}}{\tens{m}_I}.  
\end{equation}
\begin{equation}
  \label{eq:Zhang-2016-UL-v}
  \vect{v}_p^{k+1} = \vect{v}_p^{k} + \Delta t\
  \sum^{N_n}_{I}\frac{N_{Ip}^{k}\ \vec{f}_{I}^{k}}{\tens{m}_I},
\end{equation}
Later Tran \& Solowski (2019)\cite{Tran2019e} presented a
generalized-$\alpha$ scheme for the MPM inspired in the explicit time
integration algorithm proposed by Chung \& Hulbert
(1993)\cite{Geranlized_alpha_1993}, but with the particularity that
the acceleration is evaluated both in the beginning and the end of the
time step.
\begin{equation}
  \label{eq:Tran-2019-GA-v}
  \vect{v}_p^{k+1} = \vect{v}_p^{k} + \Delta t\
  \sum^{N_n}_{I} N_{Ip}^{k}\ \left[(1 - \gamma)\ \vect{a}_I^{k} +
    \gamma\ \vect{a}_I^{k+1} \right],\\
\end{equation}
\begin{equation}
\label{eq:Tran-2019-GA-x}
  \vect{x}_p^{k+1} = \vect{x}_p^{k} + 
  \sum^{N_n}_{I}N_{Ip}^{k} \left[ \Delta t\ \vec{v}_{I}^{k}+ \Delta t^2\left( (\frac{1}{2} - \beta)\
    \vec{a}_{I}^{k} + \beta\ \vec{a}_{I}^{k+1} \right) \right]
\end{equation}
\begin{equation}
  \label{eq:Tran-2019-GA-a}
  \vect{a}_p^{k+1} = \sum^{N_n}_{I}N_{Ip}^{k}\ \vec{a}_{I}^{k+1}.
\end{equation}

This scheme has prof to damps out the higher frequency noises
\cite{Tran2019e}. But it can present the same numerical instabilities
as in \eqref{eq:Sulsky-1994-UL-x},\eqref{eq:Sulsky-1994-UL-v} when
nodal masses become infinitesimal.

@@

In this section, an explicit predictor-corrector time integration
scheme is presented. It is based in the Newmark a-form 
$\gamma = 0.5$ and $\beta = 0$ which is the central difference
explicit. This method is devoted to solve a system of equations of type
\begin{equation*}
  \Matrix{M}_{IJ}\ddot{\Vector{d}}_{J} + \Matrix{C}_{IJ}\dot{\Vector{d}}_{J} +
  \Matrix{K}_{IJ}\Vector{d}_{J} = \Vector{F}_{I}.
\end{equation*}

As the MPM has a nodal stage, it is possible to apply this methods
successfully in the MPM framework as was proved by
\cite{Tran2019e}. Taking the predictor definition from the classic
literature \cite{Hughes2000} and calculating nodal velocity, and
updating particles position employing nodal values of velocity and
acceleration. Therefore we get the following expression for the
\textbf{predictor} stage for both velocity and displacement.
\begin{equation}
  \label{eq:5}
  \vec{x}_p^{k+1} = \vec{x}_p^n + \Delta t\
  \underbrace{\sum^{N_n}_{I}N_{Ip}^k\ \vec{v}_{I}^{k}}_{\vec{v}_p^{k}} + \frac{1}{2}\Delta
  t^2\ \underbrace{\sum^{N_n}_{I}\frac{N_{Ip}^k\ \vec{f}_{I}^{k}}{\tens{m}_I^k}}_{\vec{a}_p^{k}}
\end{equation}
\begin{equation}
  \label{eq:Predictor-velocity}
  \vec{v}_I^{k+1} = \vec{v}_I^n + (1 - \gamma)\ \Delta t\ \vec{a}_I^k
  \quad , \quad \vec{a}_I^k = \frac{\vec{f}_{I}^{k}}{\tens{m}_I^k}
\end{equation}

Consequently, the \textbf{corrector} stage for the nodal velocity is in the
following way
\begin{equation}
  \vec{v}_{I}^{k+1} = \vec{v}_{I}^{k+1} + \gamma\ \Delta t\
  \vec{a}_I^{k+1} \quad , \quad  \vec{a}_I^{k+1} = \frac{\vec{f}_{I}^{k+1}}{\tens{m}_I^{k+1}} 
\end{equation}

Notice that the predictor of the particle displacement is
computed using information only from the predictor of the
previous step. However, velocities are computed using half information
of the past corrector and half of the past predictor. Therefore here we share
similarities with the \textbf{leapfrog integration} which updates the
position at full time step, but updates the velocity at half time
steps. Notice also that with this approach the calculation of nodal
momentum vales in not required, and particles velocities is just a
post-process. So during the lagrangian step only the particles
position is truly computed. This minimizes the information
transference between the particles and nodes, and in consequence
minimizes the associated error to this process.

Due to its simplicity allows be implemented with minor modifications
over a standard forward Euler. It is summarized in shape of
pseudo-algorithm \ref{alg:PCE-algorithm}.

\begin{algorithm}
  \label{alg:PCE-algorithm}
  \floatname{algorithm}{Algorithm}
  \renewcommand{\thealgorithm}{}
  \caption{Explicit Predictor-Corrector scheme}
  \begin{algorithmic}[1]
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% º
    \STATE \textbf{Explicit Newmark Predictor} ($\gamma = 0.5$):
    \begin{equation*}
      \vec{v}_I^{k+1} = \vec{v}_I^n + (1 - \gamma)\ \Delta t\ \vec{a}_I^k
      \quad , \quad \vec{a}_I^k = \frac{\vec{f}_{I}^{k}}{\tens{m}_I^k}      
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Impose essential boundary conditions}:\\
    At the fixed boundary, set $\vec{v}_{I}^{k+1} = 0$. 
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    % \STATE \textbf{Discard the previous nodal values}.
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Update mass matrix}:
    \begin{equation*}
      \tens{m}_{I} = \sum_{p}^{N_p}N_{Ip}^{k+1}\ m_p,
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Deformation tensor increment calculation}.
    \begin{align*}
      &\dot{\tens{\varepsilon}_{p}}^{k+1} = \frac{1}{2}\ \sum_{I}^{N_n}
        \Grad{N_{Ip}^{k+1}}\vec{v}_{I}^{k+1} +
        (\Grad{N_{Ip}^{k+1}}\vec{v}_{I}^{k+1})^T \\
      &\Delta \tens{\varepsilon}_{p}^{k+1} = \Delta t\ \dot{\tens{\varepsilon}_{p}}^{k+1}
    \end{align*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Update the density field}:
    \begin{equation*}
      \rho_p^{k+1} = \frac{\rho_p^k}{1 + \mathit{tra}\left[\Delta\tens{\varepsilon}_{p}^{k+1}\right]}.
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Balance of forces calculation}:\\
    Calculate the total grid nodal force $\vec{f}_{I}^{k+1} =
    \vec{f}_{I}^{int,k+1} + \vec{f}_{I}^{ext,k+1}$ evaluating
    \eqref{eq:nodal_internal_forces} and \eqref{eq:nodal_external_forces} in the time step $k+1$. In the
    grid node $I$ is fixed in one of the spatial dimensions, set it to
    zero to make the grid nodal acceleration zero in that direction.\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
    \STATE \textbf{Explicit Newmark Corrector} ($\gamma = 0.5$):
    \begin{equation*}
      \vec{v}_{I}^{k+1} = \vec{v}_{I}^{k+1} + \gamma\ \Delta t\
      \vec{a}_I^{k+1} \quad , \quad  \vec{a}_I^{k+1} = \frac{\vec{f}_{I}^{k+1}}{\tens{m}_I^{k+1}}  
    \end{equation*}
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \STATE \textbf{Update particles position and velocity}:
    \begin{align*}
      &\vec{x}_p^{k+1} = \vec{x}_p^n + \Delta t\
        \sum^{N_n}_{I} N_{Ip}^k\ \vec{v}_{I}^{k} +
        \frac{1}{2}\Delta t^2\ \sum^{N_n}_{I}\frac{N_{Ip}^k\
        \vec{f}_{I}^{k}}{\tens{m}_I^k}\\
      &\vec{v}_p^{k+1} = \vec{v}_p^n + \Delta t\
        \sum^{N_n}_{I}\frac{N_{Ip}^k\
        \vec{f}_{I}^{k}}{\tens{m}_I^k}
    \end{align*}
  \end{algorithmic}
\end{algorithm}

\subsection{Analysis of the Explicit Predictor-Corrector accuracy}
\label{sec:EPC-accuracy}

In this section, the benchmark proposed by Dyka \& Ingel
(1995)\cite{Dyka1995} is considered to illustrate the
capacity of the proposed time integration algorithm to avoid velocity
fields instabilities.

In the one-dimensional bar sketched in the figure \ref{fig:Dyka_Bar},
the left end of the bar is fixed and the right and an initial velocity
$v_0 = 5\ m/s$ is given to the last quarter of it in the x positive
direction. The length of it is 0.1333 meters with an unit section.

The elastic parameters consider for this test are:
\begin{itemize} 
\item  Density : $7833\ kg/m3$
\item  Poisson ratio : $0$
\item  Elastic modulus : $200 \cdot 10^9\ Pa$
\end{itemize}

The total simulation time is 0.0002 seconds, a direct consequence of
that is we let the elastic wave generated travel thorough the bar
(from A to C and back to A) at least four times. For the spatial
discretization we will set a mesh size of 3.3325 millimeters. The
initial mesh will be composed by a single particle in the center of
the element. Regarding this, here we give one consideration to
properly modellize this benchmark under the MPM framework. Notice that
the free border of the bar has a horizontal displacement of 0.033
millimeters, therefore a computational domain with an extra gap of
0.033 millimeters is required in order to accommodate the
unconstrained displacement of the particles in the left border of the
bar. Here below the time step is controlled by the courant number, in
this case, we adopt a middle value of it (CFL=0.5) as a compromise
between stability and numerical accuracy.

The analytical solution in terms of velocity
\ref{fig:vel_analytics_dyka}, and stress
\ref{fig:stress_analytics_dyka} can be found in the appendix
\ref{app:analytical_sol} given by the Method of characteristics.

\begin{figure}\sidecaption
  \centering
  \resizebox{\hsize}{!}{
    \includegraphics[width=\textwidth]{./Figures/FE_vs_PCE_CFL_05}
  }
  \caption{Velocity evolution at the point in the bar left side.}
  \label{fig:Dyka-PCE-FE}
\end{figure}

@@

For the convergence analysis, the root-mean-square (RMS) error in the
displacement is computed. RMS error is defined as follows

@@

MPM \cite{Sulsky1994} and OTM
models to avoid tensile instabilities. Here we employ our in-house
codes for both MPM and OTM simulations.
